How to improve your relationship with your future self, V 2.0

### 2016-06-27 workflow.Rmd
Jake working on Maartens edits

---
title: How to improve your relationship with your future self.
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team. Nothing in this paper reflect the official position of the US Government. Many thanks to Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn and Cara Wong for direct help on this document and to Mika LaVaque-Manty and Ben Hansen for many useful discussions on this topic. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group. Special thanks, the EGAP Learning Days 2016 participants in Santiago Chile,  to the BITSS team and the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

<!-- An example comment -->

  <!-- Separate the thanks sections from affiliation and integrate? Wasnt sure how to do this here-->
```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.

@@

require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small",    # slightly smaller font for code
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=132,
  size='footnotesize',
  out.width='.9\\textwidth',
  message=FALSE,
  comment=NA)
```

@@

options(SweaveHooks=list(fig=function(){
			   par(mar=c(3.5, 3, 1.1, 0),
			       pty="s",
			       mgp=c(1.5,0.5,0),
			       oma=c(0,0,0,0))},
			 echo=function(){options(continue=" ") ##Don't show "+" prompts,
			 options(prompt=" ")
			 }),
	digits=4,
	scipen=8,
	width=132
	)

@@

Memory is tricky. When we really want to remember something we should practice it until it becomes internalized. In fact, learning requires effort. And when we do not practice, we tend to be overconfident in our recall abilities [@koriat2005illusions] and false memories can be implanted.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] Moreover, most of us live busy lives. We write text and code behind our new standing desk at home while the kids run around, cry about an elder sister taking away a favorite new yellow school bus toy, drinks spill on the floor, the laundry needs going, the kettle boils... Our minds are full and we need to efficiently move from task to task. If we cannot count on memory, then how can we do science?

How long does it take from planning a study to publication? How much time passes to the first reproduction of a study? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, years past the initial publication of the article.^[We note that sometimes this reproduction occurs within the context of classes and learning about statistics and data analysis. Other times, this process is organized more broadly (cite IIIE program, researchers have started to implement psychological experiments using the same protocols across many locations, see Open Science Foundation 2014, 2015, within experimental economics a team did the same, see Camerer et al 2016 Science paper. Another example is to share analysis across many researchers. in a their 'crowdsourcing analysis" Silberzahn and Uhlmann (Nature 2015) asked 29 research teams to run the analysis for a project that looked at the impact of skin color on the probability of getting a foul card in a football match.<!-- Can you do the correct citing here -->]

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make such reproduction occur much more easily and quickly.  Specifically, this piece aims to amplify some of what we already ought to know and do, and highlight some current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations.] We aim to provide practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on other bad past decisions but not on decisions about data analysis and the production of scholarly papers.

<!-- JAKE: we dont say anything about pre registration, should we? -->

@@

All results (numbers, comparisons, tables, figures) should arise from
code. Not from a series of mouse clicks or copying and pasting. If you wanted
to re-create the figure from say 2011 but include a new variable or
specification, you should be able to do so with just a few edits to the
code rather than knowledge of how you used a pointing device in
your graphical user interface all those years ago.

@@

```{r eval=FALSE}
## This file produces a plot relating the explanatory variable to the outcome.
thedata <- read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf("fig1.pdf") ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2 <- please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
```

Now, in the future if you wonder how 'that plot on page 10' was
created, you will know: (1) 'that plot' is from a file called
`fig1.pdf`and (2) `fig1.pdf`was created in
`makefig1.R`. Any time in a future, provided R still exists, changing the figure
will require some quick edits of commands already written.  In a future
where R does not exist, you (or someone else) will at least be able to read the plain text
R commands and use them to write code in a new favorite statistical
computing language: R scripts are written in
[plain text](http://en.wikipedia.org/wiki/Plain_text), and plain
text is a format that will be around as long as computer programmers
write computer programs.^[Since R is open
  source, you will also be able to download an old version of R, download an
  old-fashioned open-source operating system (like Ubuntu 10), and
  run the old-fashioned statistical computing environment in the
  old-fashioned operating system in a virtual machine on my
  new-fashioned actual machine.]

Moreover, coding saves time. Often, a lot of time, especially if projects grow (in the number of files, partners, analysis). Manually importing
a data file into R, may be effortless. Importing 100 files is another issue, and the time costs of each manual action add up quickly (and the probability for error increases).

Also, realize that *file names send messages* to your future self. This means that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you will remove some regret from your future self. For example, if you are studying inequality and protest, you might try naming a file something like  `inequality-and-protest-figures.R` instead of `temp9.R` or
`supercalifragilisticexpialidocious.R`.  More on this below. By the way, the extension
`.R` tells us and the operating system that the file contains R commands. This part of the filename enables us to quickly search our
antique hard drives for files containing R scripts.

Coding helps us to avoid making mistakes. For example, in our example above we may be interested in how many people protest. We may use a data file containing all protests for several years. We can then create a table with the number of protesters by copying the results manually to a document you use to write your paper. In copying we can make mistakes. So a better approach is to ask R (or STATA, or another program of choice) to automatically create the table (in the format you like, with horizontal lines, 3 decimals, a title, etc) and output this in whatever file type you need (.pdf, .rtf, etc). Now when we obtain new data (for additional years, or updated figures for existing data) a new table can be created quickly, so we make less mistakes and save time!

<!-- Can you include some code here to provide an example of automation of results? -->

@@

Data analysis involves a long series of decisions. Each decision
requires a justification.  Some decisions will be too small and
technical for inclusion in the published article itself. Still, these need to
be documented in the code itself [@nagler1995coding]. Paragraphs
and citations in the publication will justify the most important
decisions. So, one must code to communicate with yourself and
others. There are two main ways to avoid forgetting the reasons you
did something with data: comment your code and tightly link your code
with your writing.

@@

Comments, unexecuted text inside of a script, are a message to
collaborators (including your future self) and other consumers of your
work. In the above code chunk, we used comments to explain the lines to
readers unfamiliar with R and to remember that we had tried a different
specification but decided not to use it because adding the squared
term did not really change the substantive story arising from the
model.^[R considers text marked with `#` as a comment. For STATA simply add a `*` before the text]
Messages left for your future self (or near-future others) help
retrace and justify your decisions as the work moves from seminar
paper to conference paper to poster back to paper to dissertation and
onwards.

Notice one other benefit of coding for an audience: we learn by
teaching. By assuming that others will look at your code, you will be
more likely to write clearer code, or perhaps even to think more
deeply about what you are doing as you do it.

Comment liberally. Comments are discarded when R or STATA runs analysis, or LaTeX or pandoc
turns plain text into pdf or html files so only those who dig into
your work will see them.

@@

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf file or some other format which displays page
images to ease reading rather than to encourage reanalysis and
rewriting. Eventually pieces of that report (tables, graphs,
paragraphs) ought to show up in, or at least inform, the publishable
paper. Re-creating those analyses by pointing, clicking, copying, or
pasting would invite typing error and waste time. Re-creating your
arguments justifying your analysis decisions would also waste
time.

More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the 'source code' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016 version of this paper, this idea is now widespread and made much easier than before via online services for code sharing and collaboration such as [GitHub](http://github.com), [Open Science Framework](https://osf.io/) and [BitBucket](http://bitbucket.org), see more below.]

@@

For example, in [@sec:data-analys-comp], we suggested that we know
where 'that plot on page 10' comes from by making sure we had a
`fig1.pdf` file produced from a clearly commented plain text file
called something like `makefig1.R`.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
markdown markup with R code chunks to make things like
Figure \ref{fig:giniprot}. This combination of markdown and R is called
[R Markdown](http://rmarkdown.rstudio.org).

<!-- JAKE in the example below can you also add a line of text to people see the combination of text and table? -->

@@

```{r giniprot,message=FALSE,echo=FALSE,fig=TRUE,include=TRUE,width=4,height=4,fig.cap="Average number of Protest activity by income inequality.",out.width=".5\\textwidth",fig.env="figure",fig.lp="fig:",fig.pos="H",fig.align="center"}
##  Make a scatterplot of Protest by Inequality
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
with(good.df,plot(gini04,protest05,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Proportion Reporting Protest Activities\n(World Values Survey 2005)',
                  cex=.8))
##  Label a few interesting points
with(good.df[c("Brazil","United Kingdom","United States","Sweden","Chile"),],
     text(gini04,protest05,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
```

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in LaTeX or
`{#fig:giniprot}` in the pandoc version of markdown, I do not need to keep
track of the figure number, nor do extra work when I reorganize the document in
response to reviewer suggestions. Nor do I need a separate `makefig1.R`file or
`fig1.pdf` file.  Tables and other numerical results are also possible to
generate within the source code of a scholarly paper. Those who view the code
for this essay will see how Table \ref{tab:protest} was also generated directly
from a regression object.\footnote{@beck2010reg inspired this particular
presentation of a linear model.}

For STATA users, there now is Markdoc (https://github.com/haghish/MarkDoc) which can be used similarly.

```{r maketable, echo=FALSE, results='hide'}
  lm1<-lm(protest05~gini04+meanpr,data=good.df)  #  Run the regression

##  The next lines extract relevant and useful output from the
##  regression object, following Neal Beck's TPM article 2010.
themat<-cbind(summary(lm1)$coef[,c("Estimate","Std. Error")],confint(lm1))
n<-length(lm1$fitted.values)
r2<-signif(summary(lm1)$r.squared,2)
est.sigma<-signif(summary(lm1)$sigma,2)
colnames(themat)[3:4]<-c("~{}","~{}")
row.names(themat)<-c("Intercept","Income Inequality (lower=more equal)","Mean Political Rights (lower=more rights)")

library(xtable)
thextab<-xtable(themat,align=c("l",rep(".",4)),digits=1,
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from [@norris2015data].",label="tab:protest")
```

```{r printtable,echo=FALSE,results='asis',message=FALSE}
  print(thextab,include.colnames=FALSE,
        hline.after=NULL,
        table.placement="H",
	comment=FALSE,
        add.to.row=list(
          pos=list(0,3),
          command=c(
            "\\toprule \n &
             \\multicolumn{1}{c}{Coef} &
             \\multicolumn{1}{c}{Std. Err.} &
             \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
            paste("& \\multicolumn{4}{c}{n: ",n,",",
                  "   resid.sd: ",signif(est.sigma),
                  ", R$^2$: ",r2,"} \\\\ \\bottomrule",sep="")
            )
          ))
```

<!-- I how many of our readers will know what Sweave is.. drop or change? -->

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.^[The R project has a task view devoted to
  [reproducible research](http://cran.r-project.org/web/views/ReproducibleResearch.html) listing many of the different approaches to literate
  programming for R.}  If your workflow does not involve R, you can
still implement some of the principles here. Imagine creating a style
in Microsoft Word called 'code 'which hides your code when you print
your document, but which allows you to at least run each code chunk
piece by piece (or perhaps there are ways to extract all text of style
code from a Microsoft Word document). Or imagine just using some
other kind of indication linking paragraphs to specific places in code
files. There are many ways that creative people can program in a
literate way.

Literate programming need not go against the principle of modular data
analysis [@nagler1995coding]. Jake routinely uses
several different Sweave files that fulfill different functions, some
of them create \LaTeX code that he can `\input` into his
`main.tex` file, others setup the data, run simulations, or allow
him to record his journeys down blind alleys. Of course, when we have
flying cars running on autopilot, perhaps something other than
Sweave will make our lives even easier. Then we'll change.

**Step 2**
We analyze data in order to explain something about the world to other
scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job
with the data analysis itself: we will learn as we focus on teaching,
and we will avoid errors and save time as we ensure that others
(including our future selves) can retrace our steps. A document than
can be 'run' to reproduce all of the analyses also instills
confidence in readers and can more effectively spur discussion and
learning and cumulation of research.

@@

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on ('load'/ 'open '/ 'get '/
'use') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
`load("thedata.rda")`or `read.csv("thedata.csv")`or some
equivalent line in them, and `thedata.csv`should be stored in
some place easy to find (like in the same directory as the file or
perhaps in `"Data/thedata.rda"`). Of course, its even better to
include a comment pointing to the data file.

Where should one store data files? An obvious solution is always to
make sure that the data file used by a command file is in the same
directory as the command file. More elegant solutions require all
co-authors to have the same directory structure so that
`load("Data/thedata.rda")` means the same thing on all computers
used to work on the project. This kind of solution is one of the
things that Dropbox and more formal version control systems do well (as discussed
a bit more below in [@sec:vers-contr-prev]).

>>>>

---
title: How to improve your relationship with your future self, V 2.0
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group.  \emph{Acknowledgements:} Many thanks to the EGAP Learning Days 2016 participants in Santiago Chile, to the BITSS team and the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001. The previous version of this paper benefited from comments and discussions with Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn, Cara Wong, Mika LaVaque-Manty and Ben Hansen. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

<!-- An example comment -->
<!-- Separate the thanks sections from affiliation and integrate? Wasnt sure how to do this here-->

```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.

@@

require(knitr)
opts_chunk$set(
  tidy=TRUE,     # display code as typed
  size="footnotesize",    # slightly smaller font for code
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=100,
  out.width='.9\\textwidth',
  message=FALSE,
  comment=NA)
```

@@

options(SweaveHooks=list(fig=function(){
			   par(mar=c(3.5, 3, 1.1, 0),
			       pty="s",
			       mgp=c(1.5,0.5,0),
			       oma=c(0,0,0,0))},
			 echo=function(){options(continue=" ") ##Don't show "+" prompts,
			 options(prompt=" ")
			 }),
	digits=4,
	scipen=8,
	width=100
	)

@@

Memory is tricky. Learning requires effort. When we do not practice and repeat something that we want to remember, most people forget quickly, are overconfident in their abilities to recall future information [@koriat2005illusions], and may even recall events that never happened.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] Moreover, most of us live busy lives. We write text and code behind our new standing desk at home while the kids run around, cry about an elder sister taking away a favorite new yellow school bus toy, drinks spill on the floor, the laundry needs doing, the kettle boils \ldots Our lives and minds are full and we need to efficiently move from task to task. If we cannot count on memory, then how can we do science?

How long does it take from planning a study to publication? How much time passes to the first reproduction of a study? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, many years past the initial publication of the article.^[The process of reproducing past findings can occur when one researcher wants to build on the work of another. It can also occur within the context of classes --- some professors assign reproduction tasks to students to aid learning about data analysis and statistics. In addition to those models, recently reproduction of research has been organized to enhance the quality of public policy in the field of economic development by the 3ie Replication Program (@brown2014quality) and to assess the quality of scientific research at the level of subdisciplines in social psychology (@open2015estimating) and within experimental economics (@camerer2016evaluating). Another of 29 research teams recently collaborated on a project focusing on applied statistics to see if the same answers would emerge from re-analyses of the same data set (@silberzahn2015crowdsourced).]

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make reproducible data analysis easy and quick. Specifically, this piece aims to amplify some of what we already ought to know and do, and highlight some current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations, <http://osf.io> and <http://www.bitss.org/>.] We aim to provide practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on bad past decisions that do not have to do with data analysis and the production of scholarly papers.

@@

All results (numbers, comparisons, tables, plots, figures) should arise from
code not from a series of mouse clicks or copying and pasting.^[To our future
selves: both of us are using a computer control device called a track-pad, and
haven't used an older device called a
[mouse](https://en.wikipedia.org/wiki/Computer_mouse) in some years. We no
longer hear a clicking sound when we operate the control device either. We are
not sure why we talked about mouse clicks just now.] If you wanted to re-create
the figure from say 2011 but include a new variable or specification, you
should be able to do so with just a few edits to the code rather than knowledge
of how you used a pointing device in your graphical user interface all those
years ago.

@@

```{r eval=FALSE,tidy=FALSE}
## This file produces a plot relating the explanatory variable to the outcome.
thedata <- read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf("fig1.pdf") ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2 <- please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
```

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf`and (2)
`fig1.pdf`was created in `makefig1.R`. Any time in a future, provided R still
exists, changing the figure will require some quick edits of commands already
written.^[We use data from @norris2015data throughout this paper.]  In a future where R does not exist, you (or someone else) will at
least be able to read the plain text R commands and use them to write code in a
new favorite statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.^[Since
R is open source, you will also be able to download an old version of R,
download an old-fashioned open-source operating system (like Ubuntu 22), and
run the old-fashioned statistical computing environment in the old-fashioned
operating system in a virtual machine on your new-fashioned actual machine.]

Moreover, coding saves time. Often, a lot of time, especially if projects grow
in the number of files, partners, or complexity of analysis. Manually importing
one data file into R, may be effortless. Importing 100 files is another issue,
and the time costs of each manual action add up quickly (and the probability
for error increases).

Also, realize that *file names send messages* to your future self. This means
that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you
will remove some regret from your future self and protect your friendships and
working relationships. For example, if you are studying inequality and protest,
you might try naming a file something like `inequality-and-protest-figures.R`
instead of `temp9.R` or `supercalifragilisticexpialidocious.R`.  By the way,
the extension `.R` tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our antique
hard drives for files containing R scripts.^[We think that some method of
tagging files by the purpose of the file will continue help analysts find and organize their files for long after the idea of a computer mouse ceases to make sense.]

Coding helps us avoid making mistakes. For example, in our example above we may
be interested in how many people protest. We may use a data file containing all
protests for several years. People often create a tabular display of this data
by copying the results manually to the working paper document. In copying we
can make mistakes. So a better approach is to ask your favorite data analysis
program to automatically create the table (in the format you like, with
horizontal lines, 3 decimals, a title, etc) and output this in whatever file
type you need (.pdf, .rtf, etc). Now when we obtain new data (for additional
years, or updated figures for existing data) a new table can be created
quickly, so we make less mistakes and save time!

For example, the following table was created entirely with code using R and the `xtable` package @dahl2016aa.\footnote{@beck2010reg inspired this particular
presentation of a linear model.}

```{r loadlibraries, echo=FALSE}
makebecktable<-function(lmobj,vars,thecaption,thelabel,filename){
	require(xtable)
	##  The next lines extract relevant and useful output from the
	##  regression object, following Neal Beck's TPM article 2010.
	themat<-cbind(summary(lmobj)$coef[,c("Estimate","Std. Error")],confint(lmobj))
	n<-length(lmobj$fitted.values)
	r2<-signif(summary(lmobj)$r.squared,2)
	est.sigma<-signif(summary(lmobj)$sigma,2)
	colnames(themat)[3:4]<-c("~{}","~{}")
	row.names(themat)<-vars
	thextab<-xtable(themat,align=c("l",rep(".",4)),digits=1,
			caption=thecaption,
			label=thelabel)
	print(thextab,
	      file=filename,
	      include.colnames=FALSE,
	      hline.after=NULL,
	      table.placement="H",
	      comment=FALSE,
	      sanitize.text.function=function(x){x},
	      add.to.row=list(
			      pos=list(0,3),
			      command=c(
					"\\toprule \n &
					\\multicolumn{1}{c}{Coef} &
					\\multicolumn{1}{c}{Std. Err.} &
					\\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
				paste("& \\multicolumn{4}{c}{n: ",n,",",
				      "   resid.sd: ",signif(est.sigma),
				      ", R$^2$: ",r2,"} \\\\ \\bottomrule",sep="")
				)
			      ))

}

```

```{r maketable1, echo=TRUE, tidy=FALSE}
lm1<-lm(protest05~gini04+meanpr,data=good.df)  #  Run the regression
makebecktable(lmobj=lm1, # make the table file
	      vars=c("Intercept","Income Inequality (lower=more equal)",
		     "Mean Political Rights (lower=more rights)"),
	      thecaption="People living in countries with unequal
	      income distributions report more protest activity to World Values
	      Survey interviewers than people living in countries with relatively
	      more equal income distributions, adjusting for average political
	      rights as measured by Freedom House 1980--2010.",
	      thelabel="tab:protest",
	      filename="protesttable.tex")
```

```
\input{protesttable.tex} ## read in the table file
```

\input{protesttable.tex}

@@

Data analysis involves a long series of decisions. Each decision requires a
justification.  Some decisions will be too small and technical for inclusion in
the published article itself. Still, these need to be documented in the code
itself [@nagler1995coding]. Paragraphs and citations in the publication will
justify the most important decisions but the code itself documents the smaller
but still consequential decisions. So, one must code to communicate with
yourself and others. There are two main ways to avoid forgetting the reasons
you did something with data: comment your code and tightly link your code with
your writing --- making your code literate.

@@

Comments, unexecuted text inside of a script, are a message to collaborators
(including your future self) and other consumers of your work. In the above
code chunk, we used comments to explain the lines to readers unfamiliar with R
and to remember that we had tried a different specification but decided not to
use it because adding the squared term did not really change the substantive
story arising from the model.^[R considers text marked with `#` as a comment.
For Stata simply add a `*` before the text.] Messages left for your future self
(or near-future others) help retrace and justify your decisions as the work
moves from seminar paper to conference paper to poster back to paper to
dissertation and onwards maybe even to publication.

Notice one other benefit of coding for an audience: we learn by teaching. By
assuming that others will look at your code, you will be more likely to write
clearer code, or perhaps even to think more deeply about what you are doing as
you do it since you are explaining even as you write.

Comment liberally. Comments are discarded when R or STATA runs analysis, so
only those who dig into the source code of your work will see them.

@@

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf or html file or some other format which displays page
images to ease reading rather than to encourage reanalysis and
rewriting. Eventually pieces of that report (tables, graphs,
paragraphs) ought to show up in, or at least inform, the publishable
paper. Re-creating those analyses by pointing, clicking, copying, or
pasting would invite typing error and waste time. Re-creating your
arguments justifying your analysis decisions would also waste
time.

More importantly, we and others want to know why we did what we did. Such
explanations may not be very clear if we have some pages of printed code in one
hand and a manuscript in the other. Keep in mind the distinction between the
'source code' of a document (i.e. what computation was required to produce it)
and the visible, type-set page image. Page images are great for reading, but
not great for reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016
version of this paper, this idea is now widespread and made much easier than
before via online services for code sharing and collaboration such as
[GitHub](http://github.com), [Open Science Framework](https://osf.io/) and
[BitBucket](http://bitbucket.org). As more data analysis moves online, it will
become easier for cross-platform and geographically distant collaboration to
occur, see for only one set of examples, [Docker](http://docker.com) or other
services that make what is currently called 'cloud-computing' easier and more
accessible.]

@@

For example, in [@sec:data-analys-comp], we suggested that we knew where 'that
plot on page 10' comes from by making sure we had a `fig1.pdf` file produced
from a clearly commented plain text file called something like `makefig1.R`.
An even easier solution would be to directly include a chunk of code to produce
the figure inside of the paper itself. This paper, for example, was written in
plain text using markdown markup with R code chunks to make things like Figure
\ref{fig:giniprot}. This combination of markdown and R is called [R
Markdown](http://rmarkdown.rstudio.org). Here, is an example of embedding a
plot inside of a document. In a document sent to a journal one would tend to hide the code chunks and thus one would only see the plot.

@@

```{r giniprot,message=FALSE,echo=FALSE,fig=TRUE,include=TRUE,width=4,height=4,fig.cap="Average number of protest activities by income inequality across countries in 2004--2005.",out.width=".5\\textwidth",fig.env="figure",fig.lp="fig:",fig.pos="H",fig.align="center"}
##  Make a scatterplot of Protest by Inequality
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
with(good.df,plot(gini04,protest05,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Proportion Reporting Protest Activities\n(World Values Survey 2005)',
                  cex=.8))
##  Label a few interesting points
with(good.df[c("Brazil","United Kingdom","United States","Sweden","Chile"),],
     text(gini04,protest05,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
```

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in \LaTeX or
`{#fig:giniprot}` in the pandoc version of markdown, we do not need to keep
track of the figure number, nor do extra work when we reorganize the document
in response to reviewer suggestions. Nor do we need a separate `makefig1.R`file
or `fig1.pdf` file.  Tables and other numerical results are also possible to
generate within the source code of a scholarly paper. For example, if we had
ommitted the filename in `makebecktable` above, we would have seen a LaTeX
table be generated within this document itself. For Stata users, there now is
[Markdoc](https://github.com/haghish/MarkDoc) which is very similar to R
markdown.

The R project has a task view devoted to [reproducible
research](http://cran.r-project.org/web/views/ReproducibleResearch.html)
listing many of the different approaches to literate programming for R.  If
your workflow does not involve R, you can still implement some of the
principles here. Imagine creating a style in Microsoft Word called 'code 'which
hides your code when you print your document, but which allows you to at least
run each code chunk piece by piece (or perhaps there are ways to extract all
text of style code from a Microsoft Word document). Or imagine just using some
other kind of indication linking paragraphs to specific places in code files.
There are many ways that creative people can program in a literate way.

Literate programming need not go against the principle of modular data analysis
[@nagler1995coding]. Jake routinely uses several different files that fulfill
different functions, some of them create \LaTeX code that he can `\input` into
his `main.tex` file, others setup the data, run simulations, or allow him to
record his journeys down blind alleys. Of course, when we have flying cars
running on autopilot, perhaps something other than combining R and Markdown or
LaTeX will make our lives even easier. Then we'll change.

**Step 2** We analyze data in order to explain something about the world to
other scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job with the data analysis
itself: we will learn as we focus on teaching others about why we did what we
did, and we will avoid errors and save time as we ensure that others (including
our future selves) can retrace our steps. A document than can be 'run' to
reproduce all of the analyses also instills confidence in readers and can more
effectively spur discussion and learning and cumulation of research.

@@

All files containing commands operating on data must refer to a data file. A
reference to a data file is a line of code the analysis program will use to
operate on ('load'/ 'open '/ 'get '/ 'use') the data file. One should not have
to edit this line on different computers or platforms in order to execute this
command.  Using R, for example, all analysis files should have
`load("thedata.rda")`or `read.csv("thedata.csv")`or some equivalent line in
them, and `thedata.csv`should be stored in some easy to find place (like in the
same directory as the file or perhaps in `"Data/thedata.rda"`). Of course, its
even better to include a comment pointing to the data file in addition to the
line loading the file itself.

Where should one store data files? An obvious solution is to
make sure that the data file used by a command file is in the same
directory as the command file. More elegant solutions require all
co-authors to have the same directory structure so that
`load("Data/thedata.rda")` means the same thing on all computers
used to work on the project. This kind of solution is one of the
things that Dropbox and more formal version control systems do well (as discussed
a bit more below in [@sec:vers-contr-prev]).

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
[@nagler1995coding]. So, perhaps your analysis oriented files will
`load("cleandata.rda")`and a comment in the code will alert the future you
(among others) that `cleandata.rda`was created from `create-cleandata.R` which
in turn begins with
`read.csv(\url("http://www.greatfreedata.gov/dirtydata.csv"))`. Such a data
processing file will typically end with something like
`save("cleandata.rda")`so that we are doubly certain about the provenance of
the data.^[Of course, if you need math or paragraphs to explain what is
happening in these files, you might prefer to make them into R+Markdown or
R+LaTeX files, for which the conventional extension is .Rmd or .Rnw
respectively. So you'd have `create-cleandata.Rnw` written as a mixture of
LaTeX and R which might explain and explore the different coding decisions you
made, perhaps involving some diagnostic plots.]
